Recent advances in random matrix theory applied to heavy-tailed multivariate time series

### 2015-08-12 papers/Number1/paper.tex
Corrected some typos.

\subsection{Overview of light-tailed case}\label{sec:light_tail_review}
Yin, Bai and Krishnaiah proved in their 1988 paper
\cite{yin:bai:Krishnaiah:1988} the followng result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11} = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into empirical covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is refered
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, $\forall \zeta > 0, \exists C > 0, \exists \delta >
0$,
\[
\P\left[
  \left|
    \lambda_{(j)}(S_{n,p}) - \gamma_{j}^{n, p}
  \right| > \varphi^{C} p^{-2/3} \tilde{j}^{-1/3}
\right] \leq p^{\delta} \exp(-\varphi^\zeta)
\]
where $\lambda_{(j)}(S_{n,p})$ is the $j$-th largest eigenvalue of
$S_{n,p}$ and
\begin{eqnarray*}
  \varphi &=& (\log p)^{\log \log p} \\
  \tilde j &=& \min\{(p \wedge n) + 1 - j, j\} \\
  \gamma_{j}^{n, p} &=& \text{inf}\left\{\gamma: \int_\gamma^{\lambda_\M}
  \mu_{n,p}(x) dx \leq {j \over p}\right\} \\
  \lambda_\M &=& (1 + \sqrt q)^2 \\
  \lambda_\m &=& (1 - \sqrt q)^2 \\
  \mu_{n,p}(x) &=& {1 \over 2\pi q x} \sqrt{(x - \lambda_\m)(\lambda_\M - x)}
\end{eqnarray*}

Tao and Vu \cite{tao:vu:2012} relaxed the expoential decay condition
of {\bf C0} and replaced it with a high-moment condition, and proved the
four-moment theorem, which was later extended by Wang
\cite{Wang:2012}. In their setting, a $p \times n$ matrix $\bf X$ is
said to obey condition {\bf C1} with constant $C_0$ if $\forall 1
\leq i \leq p, \forall 1 \leq j \leq n$, $\E X_{ij} = 0$, $\E
X_{ij}^2 = 1 $ and $\sup_{i,j} \E|X_{ij}|^{C_0} < C$ for some constant
$C$ independent of $p$ and $n$.

@@

\subsection{$\alpha \in (0,2)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.}
\subsection{$\alpha \in (0,2)$: Our extension of these results; only
  order statistics of $Z_{it}$'s necessary and no sums (proof by the
  technical result)}
\subsection{$\alpha \in [2,4)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.} 
\subsection{$\alpha \in [2,4)$: The problem of centering and restrictions on $p_n$}
\subsection{$\alpha \in [2,10/3)$: Our new result for the UNCENTERED
  sample covariance matrix; only order statistics of $Z_{it}$'s
  necessary and no sums (proof by the technical result)}
\subsection{$\alpha \in [10/3,4)$: A conjecture with partial proof for
  the UNCENTERED sample covariance matrix)}

@@

\subsection{Ratio of second to largest eigenvalue (to power become uniform), self-normalizing sums, ratio of largest eigenvalue to trace, simulations, figures,...}
\subsection{Functions acting on the eigenvalues}
\subsection{Figures: Ratios of consecutive eigenvalues, illustrations for financial data, ...}
\subsection{S\&P data using ranks}
For real financial data, it is unreasonable to expect all
the time series under consideration to have the same tail index. So a
preocedure of standardization has to be adopted. In this section we
use the rank transform.

@@

Figure \ref{fig:EigenRatio} shows the ratio of successive eigenvalues
of the matrix $a_{np}^{-2}RR'$, where the rows of matrix $R$ comprise
the rank-transformed sequences of a number of selected time series
in the S\&P 500 index.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_100_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_200_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_300_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_400_shown.pdf}
  }
  \caption{$\lambda_{(i+1)} / \lambda_{(i)}$ versus $i$}
  \label{fig:EigenRatio}
\end{figure}
It is seen that, as $i \to \infty$, $\lambda_{(i+1)} / \lambda_{(i)}
\to 1$. This is in fact expected based on the theory explained in the
previous sections. Suppose the sequences are iid and serial
correlations are absent, i.e. the matrix $M=H H'$ has only one eigen
value. Then it is clear ${\lambda_{(i+1)} \over
  \lambda_i}=\left({\Gamma_i \over
    \Gamma_{i+1}}\right)^{2/\alpha}=\left(1 - {E_{i+1} \over E_1 + E_2
    + \dots E_{i+1}}\right)^{2/\alpha} \to 1$, where $E_k,
k=1,2,\dots,i+1$ are iid. exponential rv.

>>>>

\subsection{Overview of light-tailed case}\label{sec:light_tail_review}
Yin, Bai and Krishnaiah proved in their 1988 paper
\cite{yin:bai:Krishnaiah:1988} the following result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11} = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into empirical covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is referred
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, $\forall \zeta > 0, \exists C > 0, \exists \delta >
0$,
\[
\P\left[
  \left|
    \lambda_{(j)}(S_{n,p}) - \gamma_{j}^{n, p}
  \right| > \varphi^{C} p^{-2/3} \tilde{j}^{-1/3}
\right] \leq p^{\delta} \exp(-\varphi^\zeta)
\]
where $\lambda_{(j)}(S_{n,p})$ is the $j$-th largest eigenvalue of
$S_{n,p}$ and
\begin{eqnarray*}
  \varphi &=& (\log p)^{\log \log p} \\
  \tilde j &=& \min\{(p \wedge n) + 1 - j, j\} \\
  \gamma_{j}^{n, p} &=& \text{inf}\left\{\gamma: \int_\gamma^{\lambda_\M}
  \mu_{n,p}(x) dx \leq {j \over p}\right\} \\
  \lambda_\M &=& (1 + \sqrt q)^2 \\
  \lambda_\m &=& (1 - \sqrt q)^2 \\
  \mu_{n,p}(x) &=& {1 \over 2\pi q x} \sqrt{(x - \lambda_\m)(\lambda_\M - x)}
\end{eqnarray*}

Tao and Vu \cite{tao:vu:2012} relaxed the exponential decay condition
of {\bf C0} and replaced it with a high-moment condition, and proved the
four-moment theorem, which was later extended by Wang
\cite{Wang:2012}. In their setting, a $p \times n$ matrix $\bf X$ is
said to obey condition {\bf C1} with constant $C_0$ if $\forall 1
\leq i \leq p, \forall 1 \leq j \leq n$, $\E X_{ij} = 0$, $\E
X_{ij}^2 = 1 $ and $\sup_{i,j} \E|X_{ij}|^{C_0} < C$ for some constant
$C$ independent of $p$ and $n$.

@@

\subsection{$\alpha \in (0,2)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.}
\subsection{$\alpha \in (0,2)$: Our extension of these results; only
  order statistics of $Z_{it}$'s necessary and no sums (proof by the
  technical result)}
\subsection{$\alpha \in [2,4)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.} 
\subsection{$\alpha \in [2,4)$: The problem of centering and restrictions on $p_n$}
\subsection{$\alpha \in [2,10/3)$: Our new result for the UN-CENTERED
  sample covariance matrix; only order statistics of $Z_{it}$'s
  necessary and no sums (proof by the technical result)}
\subsection{$\alpha \in [10/3,4)$: A conjecture with partial proof for
  the UN-CENTERED sample covariance matrix)}

@@

\subsection{Ratio of second to largest eigenvalue (to power become uniform), self-normalizing sums, ratio of largest eigenvalue to trace, simulations, figures,...}
\subsection{Functions acting on the eigenvalues}
\subsection{Figures: Ratios of consecutive eigenvalues, illustrations for financial data, ...}
\subsection{S\&P data using ranks}
For real financial data, it is unreasonable to expect all
the time series under consideration to have the same tail index. So a
procedure of standardization has to be adopted. In this section we
use the rank transform.

@@

Figure \ref{fig:EigenRatio} shows the ratio of successive eigenvalues
of the matrix $a_{np}^{-2}RR'$, where the rows of matrix $R$ comprise
the rank-transformed sequences of a number of selected time series
in the S\&P 500 index.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_100_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_200_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_300_shown.pdf}
  }
  \subfigure[]{
    \includegraphics[scale=0.40]{EigenRatioSP500_400_shown.pdf}
  }
  \caption{$\lambda_{(i+1)} / \lambda_{(i)}$ versus $i$}
  \label{fig:EigenRatio}
\end{figure}
It is seen that, as $i \to \infty$, $\lambda_{(i+1)} / \lambda_{(i)}
\to 1$. This is in fact expected based on the theory explained in the
previous sections. Suppose the sequences are iid and serial
correlations are absent, i.e. the matrix $M=H H'$ has only one
eigenvalue. Then it is clear ${\lambda_{(i+1)} \over
  \lambda_i}=\left({\Gamma_i \over
    \Gamma_{i+1}}\right)^{2/\alpha}=\left(1 - {E_{i+1} \over E_1 + E_2
    + \dots E_{i+1}}\right)^{2/\alpha} \to 1$, where $E_k,
k=1,2,\dots,i+1$ are iid. exponential rv.
